---
title: "2D chromatographic data analysis with twoDxc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2D chromatographic data analysis with twoDxc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{twoDxc}
  %\VignetteKeywords{mass spectrometry, chromatography, 2D}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette demonstrates tools provided by twoDxc for analyzing
two-dimensional (2D) chromatography/mass spectrometry (MS) data. This package 
relies on `xcms` and `CAMERA` to read and process MS data.

# Data import

The example data file used will be an LCxLC/MS run using a quadrupole-time-of-
flight (QTOF) MS. The sample analyzed was a tea leaf extraction for a
metabolomics study. The modulation time for the run was 60 seconds with an
approximately 30 seconds before 2D valve modulation began, specified as the dead
time.
```{r setup}
library(twoDxc)
#library(xcms)
library(dplyr)
#library(CAMERA)
library(future)

tea.file <- system.file('tea_data', 'tea_pool_1.mzML', package = 'twoDxc')

tea.data <- readMSData(tea.file, mode = 'onDisk')
```

# `xcms` preprocessing

2D chromatographic data can be processed the same as 1D data, except each
compound can appear in multiple consecutive modulations. Below, the base peak
chromatogram is plotted as if the data were 1D.
```{r}
tea.bpi <- chromatogram(tea.data, aggregationFun = 'max')
plot(tea.bpi)
```

The sawtooth pattern in the data is from the gradient separation performed at
fast, ultra high pressure conditions. At least two peaks from caffeine elute in
the modulations around 2700 seconds. The extracted ion chromatogram (EIC) for
caffeine can also be plotted to show multiple peaks per compound. `twoDxc` 
provides a function for calculating _m/z_ ranges given a parts-per-million (ppm)
tolerance. The MS used to acquire this data has a tolerance of 20 ppm.
```{r}
mz.range.195 <- calc.mz.Window(195.0880, 20)
eic.195 <- chromatogram(tea.data, mz = mz.range.195, rt = c(2500, 4000))
plot(eic.195)
```
Now peak detection will be performed on the data as if it were 1D.
```{r}
cwp <- CentWaveParam(snthresh = 100, peakwidth = c(6, 15),
                     prefilter = c(5, 1000))
tea.peaks <- findChromPeaks(tea.data %>%
                              filterRt(rt = c(4.8 * 60, 60 * 60)),
                            param = cwp)

pdp <- PeakDensityParam(sampleGroups = 1, minFraction = 1, bw = 3)
tea.peaks <- groupChromPeaks(tea.peaks, pdp)

feature.chroms <- featureChromatograms(tea.peaks, features = 1:4)
plot(feature.chroms)
```
Notice the same ions are found about 60 seconds apart.

# CAMERA annotation

We will now group ions into pseudospectra using the CAMERA package.
```{r}
tea.xsa <- xsAnnotate(as(tea.peaks, 'xcmsSet'))
# Group by fwhm
tea.xsaF <- groupFWHM(tea.xsa, perfwhm = 0.6)
# Label isotopes / Doesn't work for some reason
tea.xsaI <- findIsotopes(tea.xsaF, mzabs = 0.01)
# Group by correlation
tea.xsaC <- groupCorr(tea.xsaF)
# Find adducts
tea.xsaFA <- findAdducts(tea.xsaC, polarity = 'positive')
# Plot example spectrum
plotPsSpectrum(tea.xsaFA, pspec = c(1:4), maxlabel = 5)
```
Three of the first four pspectra are 60 seconds apart from each other and
exhibit similar spectra. We will use twoDxc to group these spectra together.
```{r}
tea.xsa2D <- group2D(tea.xsaFA, 60, 30)
print(tea.xsa2D@pspec2D)
```
# Multiple sample processing.
XCMS parts.
```{r}
teas.files <- dir(system.file('tea_data', package = 'twoDxc'),
                 full.names = T)[1:3]
teas.group.info <- data.frame(sample_name = sub(basename(teas.files),
                                                pattern = '.mzML',
                                                replacement = '',
                                                fixed = T),
                              sample_group = c(rep('lvl_1', 3)),
                              stringsAsFactors = F
                              )
teas.data <- readMSData(teas.files, pdata = new('NAnnotatedDataFrame',
                                                teas.group.info),
                        mode = 'onDisk')

xchr.multi <- findChromPeaks(teas.data %>%
                               filterRt(rt = c(4.8 * 60, 60 * 60)),
                             param = cwp)
pdp.multi <- PeakDensityParam(sampleGroups = teas.data$sample_group, 
                              minFraction = (2/3), bw = 3)
xchr.multi <- groupChromPeaks(xchr.multi, param = pdp.multi)
xchr.multi <- fillChromPeaks(xchr.multi)
```
CAMERA parts.
```{r}
xsa.multi <- xsAnnotate(as(xchr.multi, 'xcmsSet'))
xsaF.multi <- groupFWHM(xsa.multi, perfwhm = 0.6)
xsaI.multi <- findIsotopes(xsaF.multi)
xsaC.multi <- groupCorr(xsaI.multi)
xsaFA.multi <- findAdducts(xsaC.multi, polarity = 'positive')
plotPsSpectrum(xsaFA.multi, pspec = 1, maxlabel = 5)
```

