---
title: "2D chromatographic data analysis with `twoDxc`"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{2D chromatographic data analysis with `twoDxc`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{xcms, CAMERA, future.apply, twoDxc}
  %\VignettePackage{twoDxc}
  %\VignetteKeywords{mass spectrometry, chromatography, 2D}
bibliography: references.bib
csl: biomed-central.csl
---

```{r global_opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8, fig.height = 4.5, fig.align = 'center',
  message = FALSE
)
BiocStyle::markdown()
```

# Introduction

This vignette demonstrates tools provided by `twoDxc` for analyzing
two-dimensional (2D) chromatography/mass spectrometry (MS) data. This package 
relies on `xcms` and `CAMERA` to read and process MS data.

# Data import

The example data file used will be an LC$\times$LC/MS run using a
quadrupole-time-of-flight (QTOF) MS. The sample analyzed was a tea leaf
extraction for a metabolomics study. The modulation time for the run was 60
seconds with approximately 30 seconds before 2D valve modulation began,
specified as the dead time. Mass spectra were acquired at 2 Hz from 100 - 1700
*m/z*.
```{r setup, message = FALSE, results = 'hide'}
library(twoDxc)
library(dplyr)
library(future.apply)
library(ggplot2)

tea.file <- system.file('tea_data', 'tea_pool_1.mzML', package = 'twoDxc')

tea.data <- readMSData(tea.file, mode = 'onDisk')
```

# `xcms` preprocessing

2D chromatographic data can be processed the same as 1D data, except each
compound can appear in multiple consecutive modulations. Below, the base peak
chromatogram is plotted as if the data were 1D.
```{r bpi}
tea.bpi <- chromatogram(tea.data, aggregationFun = 'max')
plot(tea.bpi)
```

The chromatogram's sawtooth pattern arises from the gradient separations performed at
fast, ultra high pressure conditions. We can see that at least two peaks from 
caffeine elute in the modulations around 2700 seconds. The extracted ion 
chromatogram (EIC) for caffeine can also be plotted to show multiple peaks per 
compound. `twoDxc` provides a function for calculating _m/z_ ranges given a 
parts-per-million (ppm) tolerance. The MS used to acquire this data has a 
tolerance of 20 ppm. Note the retention time range in the plot is adjusted.
```{r eic}
mz.range.195 <- calc.mz.Window(195.0882, 20)
# m/z tolerance window for caffeine's accurate mass is:
print(mz.range.195)
eic.195 <- chromatogram(tea.data, mz = mz.range.195, rt = c(2500, 3000))
plot(eic.195)
```

Now peak detection will be performed on the data as if it were 1D with `xcms`.
```{r peak_finding, fig.width = 8, fig.height = 8}
cwp <- CentWaveParam(snthresh = 100, peakwidth = c(6, 15),
                     prefilter = c(5, 1000))
tea.peaks <- findChromPeaks(tea.data %>%
                              filterRt(rt = c(4.8 * 60, 60 * 60)),
                            param = cwp)

pdp <- PeakDensityParam(sampleGroups = 1, minFraction = 1, bw = 3)
tea.peaks <- groupChromPeaks(tea.peaks, pdp)

feature.chroms <- featureChromatograms(tea.peaks, features = 1:4)
par(mar = rep(2, 4))
plot(feature.chroms)
```

Notice the same ions are found about 60 seconds apart.

# CAMERA annotation

We will now group ions into pseudospectra using the CAMERA package.
```{r CAMERA_annotation}
tea.xsa <- xsAnnotate(as(tea.peaks, 'xcmsSet'))
# Group by fwhm
tea.xsaF <- groupFWHM(tea.xsa, perfwhm = 0.6)
# Label isotopes
tea.xsaI <- findIsotopes(tea.xsaF, mzabs = 0.01)
# Group by correlation
tea.xsaC <- groupCorr(tea.xsaF)
# Find adducts
tea.xsaFA <- findAdducts(tea.xsaC, polarity = 'positive')
# Plot example spectrum
plotPsSpectrum(tea.xsaFA, pspec = c(1:4), maxlabel = 5)
```
Three of the first four pspectra are 60 seconds apart from each other and
exhibit similar spectra. 

We will use `twoDxc` to group these spectra together. The
`group2D` function takes its arguments as: 1) the xsAnnotate object, 2) the
modulation time for the comprehensive 2D run, and 3) the dead time, or where the
time at which modulations start, to adjust the 2D retention time.

The algorithm groups features by their *m/z*, 1D RT, and 2D RT, with default
tolerances of 20 ppm, 3 minutes, and 5 seconds, respectively. Matching features
are grouped into one psg.2d group and their intensities summed. This way, we can
make more accurate comparisons between different phenotypes and cut down on
redundant features.
```{r 2D_grouping}
filter.Ion <- function(pspectra, ion, ppm = 20){
  ion.range <- calc.mz.Window(ion, ppm)
  filtered.pspectra <- as.data.frame(pspectra) %>%
    filter(mz > ion.range[1] & mz < ion.range[2])
  return(filtered.pspectra)
}

print(getpspectra(tea.xsaFA, grp = c(1:10)) %>%
        filter.Ion(195.0882))
tea.xsa2D <- group2D(tea.xsaFA, 60, 30)
print(tea.xsa2D@pspec2D %>%
  filter.Ion(195.0882))
```
# Multiple sample processing
Below follows the sample sample processing steps but for three replicates of
the tea extraction.

## XCMS parts.
```{r multiple_xcms}
teas.files <- dir(system.file('tea_data', package = 'twoDxc'),
                 full.names = T)[1:3]
teas.group.info <- data.frame(sample_name = sub(basename(teas.files),
                                                pattern = '.mzML',
                                                replacement = '',
                                                fixed = T),
                              sample_group = c(rep('lvl_1', 3)),
                              stringsAsFactors = F
                              )
teas.data <- readMSData(teas.files, pdata = new('NAnnotatedDataFrame',
                                                teas.group.info),
                        mode = 'onDisk')

xchr.multi <- findChromPeaks(teas.data %>%
                               filterRt(rt = c(4.8 * 60, 60 * 60)),
                             param = cwp)
pdp.multi <- PeakDensityParam(sampleGroups = teas.data$sample_group, 
                              minFraction = (2/3), bw = 3)
xchr.multi <- groupChromPeaks(xchr.multi, param = pdp.multi)
xchr.multi <- fillChromPeaks(xchr.multi)
```
## CAMERA parts.
```{r multiple_annotation}
xsa.multi <- xsAnnotate(as(xchr.multi, 'xcmsSet'))
xsaF.multi <- groupFWHM(xsa.multi, perfwhm = 0.6)
xsaI.multi <- findIsotopes(xsaF.multi)
xsaC.multi <- groupCorr(xsaI.multi)
xsaFA.multi <- findAdducts(xsaC.multi, polarity = 'positive')
plotPsSpectrum(xsaFA.multi, pspec = 1, maxlabel = 5)
```


## `group2D` on multiple samples
```{r serial_processing}
timer <- proc.time()
xsa2D.multi <- group2D(xsaFA.multi, 60, 30)
proc.time() - timer
```
# Parallelization

`group2D` can run in parallel if `future.apply` is installed. The below block
should demonstrate an improvement in processing time when data are grouped in
parallel.
```{r parallel_processing}
plan(multiprocess)
timer <- proc.time()
xsa2D.multi.p <- group2D(xsaFA.multi, 60, 30, parallelized = T)
proc.time() - timer
identical(xsa2D.multi, xsa2D.multi.p)
```

# Plotting 2D data

Visualizing 2D chromatography data is easier as a heatmap plot, in which
the x-axis shows 1D RT, the y-axis shows 2D RT, and the z-axis (color) shows ion
intensity.
```{r TIC_plot}
plot2D(tea.data, file = 1, 60, 30)
```
Because the total ion chromatogram (TIC) is a combination of intensities of all
ion channels, this information underrepresents the chemical diversity in this
metabolomics sample. Below is an example of an extracted ion chromatogram (EIC)
for ion 195, representing caffeine.
```{r EIC_plot}
plot2D(tea.data, file = 1, mod.time = 60, dead.time = 30, ion = 195.0882)
```

It may be useful to iterate through the m/z acquisition range and generate
EICs for each ion. The example code below would do so for m/z range 100 to 108. The upper range for the loop can be adjusted to the max *m/z* recorded.

```{r eic_loop}
data.mz.range <- c(round(min(tea.data@featureData@data$lowMZ)):108)
eics.2d <- lapply(data.mz.range, function(x){
  plot2D(tea.data, file = 1, mod.time = 60, dead.time = 30, ion = x,
         mz.tol = 'abs', save.output = F)
})
```

# References
